\documentclass{article}

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{pgfgantt}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{pgfkeys}
\usepackage{xcolor}

\title{The Gentrification of the San Francisco Bay Area: TEAM-17 Progress Report}
\author{Benicio Bailey, Jacob Feenstra, Diane Kim, Aryan Sarda}

\begin{document}
\maketitle

\section{Method}

\begin{itemize}
\item  Provide a detailed explanation of your current design (nearly done) and implementation (at least half done), covering both the visualization and storytelling aspects.
\item  List any significant changes, design decisions, or improvements made since the proposal. Include figures or diagrams if they help clarify your method.
\end{itemize}

\section{Evaluation Plan}

\begin{itemize}
\item  Describe how you plan to evaluate your system. 
\end{itemize}


\section{Preliminary Results}

\begin{itemize}
\item  Report any results/insights obtained so far. If no results are yet available, describe what results/insights you expect to generate and how they would support your claims.
\end{itemize}

\section{Plan of Activities}

\begin{itemize}
\item  Present your latest activity plan, using either a Gantt chart (example) or a table. For each task, include: (a) Who is responsible, (b) The start and end time (or duration), and (c) Whether the task is complete, ongoing, or upcoming. 
\end{itemize}

% making a gantt chart
The following Gantt chart illustrates the timeline of key project tasks:


\begin{ganttchart}[
    hgrid,
    vgrid,
    time slot format=isodate-yearmonth,
    time slot unit=month,
    title/.append style={draw=none, fill=gray!20},
    title label font=\bfseries,
    title label anchor/.style={},
    bar height=0.6,
    group right shift=0,
    group top shift=0.7,
    group height=.3,
    group peaks height=.2,
    bar label font=\small\bfseries,
    bar label node/.append style={align=right}
  ]{2025-06}{2025-12}

  \end{ganttchart}


\begin{center}

\begin{ganttchart}[
  hgrid,
  vgrid,
  time slot format=isodate,
  time slot unit=day,
  x unit=0.7cm,
  y unit chart=0.8cm,
  title/.style={draw=none, fill=gray!20},
  title label font=\bfseries\footnotesize,
  bar/.style={fill=blue!50},
  bar label font=\small\color{black},
  bar label node/.append style={align=center},
  bar height=0.6,
]{2025-05-21}{2025-06-05}

\gantttitlecalendar{year, month=name, day=2} \\

\ganttbar[
  progress=100
]{data preprocessing\\(Jacob)}{2025-05-21}{2025-05-22} \\

\ganttbar[
  progress=0
]{video presentations\\(all)}{2025-06-02}{2025-06-05} \\

\ganttbar[
  bar label text={Timeline Annotations (Jacob)},
  label text={Delivery}
]{Delivery}{2025-06-05}{2025-06-05} \\

\ganttbar[
  bar label text={Martini Glass Introduction (Jacob)},
  label text={Testing \& Review}
]{Testing \& Review}{2025-06-03}{2025-06-04} \\

\ganttbar[
  bar label text={Bay Area 9-County Map (Aryan)},
  label text={Testing \& Review}
]{Testing \& Review}{2025-06-03}{2025-06-04} \\

\ganttbar[
  progress=120
]{json county maps\\{Benicio}}{2025-06-03}{2025-06-04} \\

\ganttbar[
  progress=10
]{county vis dashboard\\{Diane}}{2025-05-23}{2025-05-26} \\

\ganttbar[
  progress=20
]{9 census tract heatmaps\\(Benicio)}{2025-05-23}{2025-05-30} \\

\ganttbar[
  bar label text={2010-2023 Timeline (Diane)},
  label text={Testing \& Review}
]{Testing \& Review}{2025-06-03}{2025-06-04} \\

\end{ganttchart}







\section{Team's Effort Division}

\begin{itemize}
\item Please provide a brief effort distribution statement at the end. It can be as simple as "all team members have contributed a similar amount of effort". If effort distribution is too uneven, we might consider assigning higher scores to members who have contributed more.
\end{itemize}

\nocite{*}
\bibliographystyle{alpha}
\bibliography{refs}

\end{document}